Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
adding periods after Lightning talk titles

The lightning talks were intended to give an opportunity for attendees to
quickly highlight an important issue or a potential solution.
%
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
Facilitating Research Training While Supporting Scientific Research}} Benjamin
Tovar presented a case study of ``The Cooperative Computing Lab'' at the
University of Notre Dame, which is a small group of individuals whose main tasks
are collaborating with people that have large-scale computing problems,
operating various parallel computer systems, conducting computer science
research, and developing open source software. One of the main challenges they
face is finding a balance between flexibility/training and stability/quality.
Their current solution for ensuring the latter was to add a software engineer
(the presenter) to the existing team of faculty and students, who now also
serves as a ``spring'' between flexibility and stability.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}} The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it is not well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly.
Instead, sustainability at its heart is a systemic concept and must be viewed
from a range of different dimensions including environmental, economic,
individual, social, and technical. She introduced the Karlskrona Manifesto on
Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
of related work on the topic of sustainability into a set of (mis-)perceptions
and principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
to Improve Productivity of Legacy Tools}} Abani Patra presented an example of
how the community using Titan2D, a geoflow simulation software, increased the
productivity of their tools by improving both code and data layout. The main
obstacles in this change were the non-existence of a common version control
system for the source code, coupled with multiple versions of the same code
base, the fixed format of input files, that many input values were set as
compilation flags, and that the internal data layout was not suitable for modern
technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
developers included reinforcing the code structure using multiple layers of
Python and C++ interfaces, and a redesign of the data layout to be more suitable
for modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software:
Lessons from Open Source}} Abigail Cabunoc Mayes combines two properties of open
source that, together, create a suitable habitat for sustainable software. The
first of the two properties, public, does not only mean public code. It also
includes public discussions, a public process of including contributions, and an
open license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability, and ease of understanding, but also
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with the Computational
Infrastructure for Geodynamics (CIG), a community software with a worldwide user
base. Like others, their main goals include high usability, sustainability, and
reproducibility. As a means to achieve these goals, various communication
channels have been developed, such as mailing lists, wikis, workshops,
hackathons, tutorials, and webinars. In order to contribute to the
infrastructure, codes must adhere to specified minimum standards with the desire
that all codes are working toward target standards.
%, but ideals are also given to encourage reaching the aforementioned goals. 
These include, e.g., the use of version control, certain coding styles, the
presence and nature of code tests and documentation, and certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
Hwang mentioned a variety of ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the CIG website. An analysis of
the resulting citations showed that $80\%$ of papers that use CIG codes mention
the code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better cited. In
part, this is attributed to the fact that CIG requires that donated software
provide a citable paper specified in the User Manual. The project is working on
tools and methods to generate attribution information automatically.
%This,
%in part, is attributed to suggested citation forms communicated to the
%community, and runtime tools that help collecting a relevant citation list
%automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to
other disciplines such as biology, astrophysics, and digital curation. It includes
physicists, digital librarians, as well as computer scientists, and aims to
achieve some commonality across disciplines. Examples are meta-data descriptions
of archived data, computational descriptions, descriptions of how data was
processed, questions such as whether computation replication can be automated,
and what the impact of access policies on the preservation infrastructure is.
One of the products of this effort is a suite of tools that deals with this
preservation, and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user community
from the start, and to budget that way. For the specific example of the
preservation software, this means that besides adding value to the community, it
needs to be transparent to their workflows, i.e., not requiring additional
effort to preserve ``research objects.''

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}} James
Hetherington started by listing some of the problems research software faces,
which include poor standard of verification and low levels of reuse. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who do not want to serve as programmers, and
creating synergies with research computing platforms. The success of such a
group would be measured by the output and quality of the research software.
Members could be part of Research Computing or faculty. They would not be
independent researchers, but would have to have a research background. An
attempt to form a community of RSE groups within the UK has been underway for
several years, including funding from the UK Research Council. An open question
is whether this approach can be adopted in other countries, including the~USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, proficient developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late interaction
with users. Instead, an alternative approach was proposed that, at first, skips
the development phase and repeatedly goes through requirement gathering, design
and user interaction/learning phases, and only eventually starts development
once an agreement is reached, leading to an increased user satisfaction, higher
adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
Platform for Predictive Biology: Lessons Learned on the KBase Project}} 
Dan Gunter presented experiences and lessons learned as part of KBase, an
open software and data platform for addressing the grand challenge of systems
biology: predicting and designing biological function. KBase is a unified system
that integrates data and analytical tools for comparative functional genomics of
microbes, plants, and their communities. However, it is also a collaborative
environment for sharing methods and results, and placing those results in the
context of knowledge in the field. Being a large, multi-institutional project,
one of the big challenges is to agree on standards to enable a single,
maintainable system. Working in isolation does not work (anymore) within this
field, and the community in the field also does not have standards for software
engineering. This is contrasted to computer science research, where software
engineering standards shorten design cycles, leading to more time for highly
rewarded activities like publishing, performance studies, graduating students,
or protecting ideas before publication. Instead of this more traditional
approach, KBase uses a variation on the ``Scrum'' methodology. After picking
projects and team members, four to five teams work on projects for about two
weeks before a one-week evaluation by the executive committee. Based on their
assessment new teams and projects might be chosen before iteratively restarting
the ``agile'' development cycle. This process is intentionally open and
documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects among
experimental data, research software, and publications. A current effort in the
Geosciences is the ``Geoscience Paper of the Future,'' which includes four
elements. First, it forms a modern paper including text, data, and pointers to
supplementary materials. Second, it is reproducible, including data processing,
workflow, and visualization tools. Third, it is part of open science, which
includes being publicly available under open licenses, and providing meta-data.
Last but not least, it uses digital scholarship elements like persistent
identifiers for software, data, and authors, and it cites both data and
software. OntoSoft is a tool for helping with this effort, providing software stewardship
for the Geosciences. As part of this initiative, a special issue of a journal in
Geoscience areas is planned to include only Geoscience papers of the future,
with submissions open until the end of 2015. In addition, training sessions are
provided to geoscientists to learn best practices in software and data sharing,
provenance documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
Accreditation Framework}} Neil Chue Hong presented a proposal to build a
scientific software accreditation framework. One of the aims of such a framework
would be to measure how ``good'' a given piece of software is, and to evaluate
how this can be effectively measured in the first place. This can be compared to
the effort of the standardized, easy to read, and understandable nutritional
labeling of food, which only contain a small set of categories. However, such a
framework for software would be more difficult due to different existing
community norms. The challenges such a framework faces include that many
measurements are subjective, that many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
Support for Sustainable Scientific Software}} Carver argued that for scientific
software to be truly sustainable, there is a need for developers to use
appropriate software engineering practices. His experience interacting with
scientific teams indicates that choosing and tailoring these practices is not a
trivial exercise. There is a general culture clash between software engineering
and science that hinders our ability to communicate and choose appropriate
methods. In addition, many experienced scientific software developers appear to
be unaware of software engineering practices that may be beneficial to them. The
most appropriate software engineering practices are those that are lightweight,
properly tailored, and focus on the key software development problems faced by
scientists. In order to increase the use of software engineering in science, we
need more documented success stories. These successes need to be socialized 
%\choinote{not sure if ``socialize'' can be used in such a way}
within the scientific community through workshops like the Software Engineering
for Science workshop
series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
and the new Software Engineering track in \textit{Computing in Science and
Engineering} magazine.

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
This talk highlighted an attempt to solve the common problem for open source
development: it is difficult to collect information about how many people use
particular software, how often, which version, which parts of the software, or
on which operating system. Current solutions include surveys, but these have
high uncertainties. A different approach is based on automatic ``call-backs''
that collect these information at runtime and send it to a central place for
analysis. Problems with this approach include obtaining agreement from the user,
legal issues, increased maintenance (of servers), ethical questions, and also the
lack of a common infrastructure. Some of these problems are of a social nature
and have to be solved as such, but the last problem (a missing common
infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
which is developing common APIs and a library implementation for common,
repeating tasks such as obtaining user consent. Results are uploaded not
directly to project servers, but to neutral third parties that only publish
aggregated statistics to projects.

\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
included)}} Alice Allen described the Astrophysics Source Code Library
(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
software that is indexed by indexing services and can be cited. ASCL offers
clones of existing infrastructure, provides server space and computing
resources, shares innovations, and permits moves elsewhere. Users provide a
domain name, then control and configure their site and use it as intended,
gather their codes as they wish, share innovations, and protect the provided
computing environment.

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
Developing Community}} The effort behind this talk given by Ray Idaszak started
from a breakout group at an NSF SI2 workshop in 2015, and centers around
building a framework for creating metrics for scientific software. This
framework would improve both the metrics and the software it evaluates, and
could also serve as a tool for building a community around the idea. With
especially the last-mentioned idea (building a community) in mind, a software
``peer review group'' would be created, representing stakeholders who will
self-review software created by their respective communities, and will
concurrently develop metrics. The whole project should be community-governed,
without a single institution overseeing the activities or infrastructure, with
the hope to evolving community-generated and adopted standards. The generation
of metrics would be tied to the actual evaluation of software, creating an
incentive by improving the evaluated software itself during this process. The
framework code would provide infrastructure for the creation of metrics and
evaluation, and forums for generation of software success metrics. It would also
support code reviews of the evaluated software. An open question is whether it
is possible to fit the resulting metrics in a common template. So far, this is
still in a design phase, with a white paper at the CSESSP workshop 2015 and this
talk, but the WSSSPE workshops are seen as a forum for the community to assemble
and act, and is planned to be used also in the future to build this community
and framework.

>>>>

The lightning talks were intended to give an opportunity for attendees to
quickly highlight an important issue or a potential solution.
%
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
Facilitating Research Training While Supporting Scientific Research}}. Benjamin
Tovar presented a case study of ``The Cooperative Computing Lab'' at the
University of Notre Dame, which is a small group of individuals whose main tasks
are collaborating with people that have large-scale computing problems,
operating various parallel computer systems, conducting computer science
research, and developing open source software. One of the main challenges they
face is finding a balance between flexibility/training and stability/quality.
Their current solution for ensuring the latter was to add a software engineer
(the presenter) to the existing team of faculty and students, who now also
serves as a ``spring'' between flexibility and stability.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}}. The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it is not well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly.
Instead, sustainability at its heart is a systemic concept and must be viewed
from a range of different dimensions including environmental, economic,
individual, social, and technical. She introduced the Karlskrona Manifesto on
Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
of related work on the topic of sustainability into a set of (mis-)perceptions
and principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
to Improve Productivity of Legacy Tools}}. Abani Patra presented an example of
how the community using Titan2D, a geoflow simulation software, increased the
productivity of their tools by improving both code and data layout. The main
obstacles in this change were the non-existence of a common version control
system for the source code, coupled with multiple versions of the same code
base, the fixed format of input files, that many input values were set as
compilation flags, and that the internal data layout was not suitable for modern
technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
developers included reinforcing the code structure using multiple layers of
Python and C++ interfaces, and a redesign of the data layout to be more suitable
for modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software:
Lessons from Open Source}}. Abigail Cabunoc Mayes combines two properties of open
source that, together, create a suitable habitat for sustainable software. The
first of the two properties, public, does not only mean public code. It also
includes public discussions, a public process of including contributions, and an
open license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability, and ease of understanding, but also
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}}. Lorraine Hwang presented experiences with the Computational
Infrastructure for Geodynamics (CIG), a community software with a worldwide user
base. Like others, their main goals include high usability, sustainability, and
reproducibility. As a means to achieve these goals, various communication
channels have been developed, such as mailing lists, wikis, workshops,
hackathons, tutorials, and webinars. In order to contribute to the
infrastructure, codes must adhere to specified minimum standards with the desire
that all codes are working toward target standards.
%, but ideals are also given to encourage reaching the aforementioned goals. 
These include, e.g., the use of version control, certain coding styles, the
presence and nature of code tests and documentation, and certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}}. In a second talk, Lorraine
Hwang mentioned a variety of ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the CIG website. An analysis of
the resulting citations showed that $80\%$ of papers that use CIG codes mention
the code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better cited. In
part, this is attributed to the fact that CIG requires that donated software
provide a citable paper specified in the User Manual. The project is working on
tools and methods to generate attribution information automatically.
%This,
%in part, is attributed to suggested citation forms communicated to the
%community, and runtime tools that help collecting a relevant citation list
%automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}}. DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to
other disciplines such as biology, astrophysics, and digital curation. It includes
physicists, digital librarians, as well as computer scientists, and aims to
achieve some commonality across disciplines. Examples are meta-data descriptions
of archived data, computational descriptions, descriptions of how data was
processed, questions such as whether computation replication can be automated,
and what the impact of access policies on the preservation infrastructure is.
One of the products of this effort is a suite of tools that deals with this
preservation, and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user community
from the start, and to budget that way. For the specific example of the
preservation software, this means that besides adding value to the community, it
needs to be transparent to their workflows, i.e., not requiring additional
effort to preserve ``research objects.''

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}}. James
Hetherington started by listing some of the problems research software faces,
which include poor standard of verification and low levels of reuse. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who do not want to serve as programmers, and
creating synergies with research computing platforms. The success of such a
group would be measured by the output and quality of the research software.
Members could be part of Research Computing or faculty. They would not be
independent researchers, but would have to have a research background. An
attempt to form a community of RSE groups within the UK has been underway for
several years, including funding from the UK Research Council. An open question
is whether this approach can be adopted in other countries, including the~USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}}. Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, proficient developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late interaction
with users. Instead, an alternative approach was proposed that, at first, skips
the development phase and repeatedly goes through requirement gathering, design
and user interaction/learning phases, and only eventually starts development
once an agreement is reached, leading to an increased user satisfaction, higher
adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
Platform for Predictive Biology: Lessons Learned on the KBase Project}}. 
Dan Gunter presented experiences and lessons learned as part of KBase, an
open software and data platform for addressing the grand challenge of systems
biology: predicting and designing biological function. KBase is a unified system
that integrates data and analytical tools for comparative functional genomics of
microbes, plants, and their communities. However, it is also a collaborative
environment for sharing methods and results, and placing those results in the
context of knowledge in the field. Being a large, multi-institutional project,
one of the big challenges is to agree on standards to enable a single,
maintainable system. Working in isolation does not work (anymore) within this
field, and the community in the field also does not have standards for software
engineering. This is contrasted to computer science research, where software
engineering standards shorten design cycles, leading to more time for highly
rewarded activities like publishing, performance studies, graduating students,
or protecting ideas before publication. Instead of this more traditional
approach, KBase uses a variation on the ``Scrum'' methodology. After picking
projects and team members, four to five teams work on projects for about two
weeks before a one-week evaluation by the executive committee. Based on their
assessment new teams and projects might be chosen before iteratively restarting
the ``agile'' development cycle. This process is intentionally open and
documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}}. Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects among
experimental data, research software, and publications. A current effort in the
Geosciences is the ``Geoscience Paper of the Future,'' which includes four
elements. First, it forms a modern paper including text, data, and pointers to
supplementary materials. Second, it is reproducible, including data processing,
workflow, and visualization tools. Third, it is part of open science, which
includes being publicly available under open licenses, and providing meta-data.
Last but not least, it uses digital scholarship elements like persistent
identifiers for software, data, and authors, and it cites both data and
software. OntoSoft is a tool for helping with this effort, providing software stewardship
for the Geosciences. As part of this initiative, a special issue of a journal in
Geoscience areas is planned to include only Geoscience papers of the future,
with submissions open until the end of 2015. In addition, training sessions are
provided to geoscientists to learn best practices in software and data sharing,
provenance documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
Accreditation Framework}}. Neil Chue Hong presented a proposal to build a
scientific software accreditation framework. One of the aims of such a framework
would be to measure how ``good'' a given piece of software is, and to evaluate
how this can be effectively measured in the first place. This can be compared to
the effort of the standardized, easy to read, and understandable nutritional
labeling of food, which only contain a small set of categories. However, such a
framework for software would be more difficult due to different existing
community norms. The challenges such a framework faces include that many
measurements are subjective, that many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
Support for Sustainable Scientific Software}}. Carver argued that for scientific
software to be truly sustainable, there is a need for developers to use
appropriate software engineering practices. His experience interacting with
scientific teams indicates that choosing and tailoring these practices is not a
trivial exercise. There is a general culture clash between software engineering
and science that hinders our ability to communicate and choose appropriate
methods. In addition, many experienced scientific software developers appear to
be unaware of software engineering practices that may be beneficial to them. The
most appropriate software engineering practices are those that are lightweight,
properly tailored, and focus on the key software development problems faced by
scientists. In order to increase the use of software engineering in science, we
need more documented success stories. These successes need to be socialized 
%\choinote{not sure if ``socialize'' can be used in such a way}
within the scientific community through workshops like the Software Engineering
for Science workshop
series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
and the new Software Engineering track in \textit{Computing in Science and
Engineering} magazine.

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}.
This talk highlighted an attempt to solve the common problem for open source
development: it is difficult to collect information about how many people use
particular software, how often, which version, which parts of the software, or
on which operating system. Current solutions include surveys, but these have
high uncertainties. A different approach is based on automatic ``call-backs''
that collect these information at runtime and send it to a central place for
analysis. Problems with this approach include obtaining agreement from the user,
legal issues, increased maintenance (of servers), ethical questions, and also the
lack of a common infrastructure. Some of these problems are of a social nature
and have to be solved as such, but the last problem (a missing common
infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
which is developing common APIs and a library implementation for common,
repeating tasks such as obtaining user consent. Results are uploaded not
directly to project servers, but to neutral third parties that only publish
aggregated statistics to projects.

\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
included)}}. Alice Allen described the Astrophysics Source Code Library
(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
software that is indexed by indexing services and can be cited. ASCL offers
clones of existing infrastructure, provides server space and computing
resources, shares innovations, and permits moves elsewhere. Users provide a
domain name, then control and configure their site and use it as intended,
gather their codes as they wish, share innovations, and protect the provided
computing environment.

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
Developing Community}}. The effort behind this talk given by Ray Idaszak started
from a breakout group at an NSF SI2 workshop in 2015, and centers around
building a framework for creating metrics for scientific software. This
framework would improve both the metrics and the software it evaluates, and
could also serve as a tool for building a community around the idea. With
especially the last-mentioned idea (building a community) in mind, a software
``peer review group'' would be created, representing stakeholders who will
self-review software created by their respective communities, and will
concurrently develop metrics. The whole project should be community-governed,
without a single institution overseeing the activities or infrastructure, with
the hope to evolving community-generated and adopted standards. The generation
of metrics would be tied to the actual evaluation of software, creating an
incentive by improving the evaluated software itself during this process. The
framework code would provide infrastructure for the creation of metrics and
evaluation, and forums for generation of software success metrics. It would also
support code reviews of the evaluated software. An open question is whether it
is possible to fit the resulting metrics in a common template. So far, this is
still in a design phase, with a white paper at the CSESSP workshop 2015 and this
talk, but the WSSSPE workshops are seen as a forum for the community to assemble
and act, and is planned to be used also in the future to build this community
and framework.
